---
title: "ETC5521: Exploratory Data Analysis"
subtitle: "Exploring bivariate dependencies, linearising"
author: "Di Cook"
email: "ETC5521.Clayton-x@monash.edu"
date: "Week 5 - Session 2"
length: "10 minutes"
color_theme: "yellow"
pdflink: "lecture-05B.pdf"
bgimg: "images/wk5bg1.jpg"
output:
  xaringan::moon_reader:
    css:
      - ninjutsu 
      - "assets/animate.css"
      - "assets/fira-code.css"
      - "assets/boxes.css"
      - "assets/styles.css"
      - "assets/custom.css"
      - "assets/monash-logo.css"
      - "assets/monash-brand.css"
      - "assets/monash-fonts.css"
      - "assets/table.css"
      - "assets/panelset.css"
      - "assets/di2.css"
    self_contained: false 
    seal: false 
    chakra: 'libs/remark-latest.min.js'
    lib_dir: libs
    includes:
      in_header: "assets/custom.html"
    mathjax: "assets/mathjax-local/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    nature:
      highlightStyle: googlecode
      highlightLanguage: r 
      highlightLines: true
      highlightSpans: false 
      countIncrementalSlides: false
      slideNumberFormat: '%current%/%total%'
      navigation:
        scroll: false 
        touch: true
        click: false
      ratio: '16:9'
---

```{r, include = FALSE}
current_file <- knitr::current_input()
```
```{r, include = FALSE, eval = F}
input <- fs::path_ext_set(current_file, "html")
pagedown::chrome_print(input = input, format = "pdf", wait = 20)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "images/week5/",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  out.width = "100%",
  fig.retina = 3,
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)
```

```{r}
library(tidyverse)
library(gt)
library(forcats)
library(gridExtra)
library(ggthemes)
library(mvtnorm)
library(ggplot2movies)
library(ggExtra)
library(housingData)
library(nullabor)
```

```{r titleslide, child="components/titleslide.Rmd"}
```

---
# History

<br>
<br>

> Scatter plots are glorious. Of all the major chart types, they are by far the most powerful. They allow us to .monash-orange2[quickly understand relationships] that would be nearly impossible to recognize in a table or a different type of chart. ... Michael Friendly and Daniel Denis, psychologists and historians of graphics, call the scatter plot the most "generally useful invention in the history of statistical graphics." [Dan Kopf](https://qz.com/1235712/the-origins-of-the-scatter-plot-data-visualizations-greatest-invention/)

---
# History

- Descartes provided the Cartesian coordinate system in the 17th century, with perpendicular lines indicating two axes.
--

- It wasn't until 1832 that the scatterplot appeared, when [John Frederick Herschel](http://www.datavis.ca/milestones/index.php?group=1800%2B) plotted position and time of double stars.
--

- This is 50 years after [bar charts and line charts](http://www.datavis.ca/milestones/index.php?group=1700s) appeared, used in the work of William Playfair to examine economic data.
--

- Kopf argues that *The scatter plot, by contrast, proved more useful for scientists*, but it clearly is useful for economics today.


.footnote[http://www.datavis.ca/milestones/]

---
class: informative middle
## Language and terminology

Are the words "correlation" and "association" interchangeable?

> .monash-gray10[In the broadest sense] **correlation** .monash-gray10[is any statistical association, though it commonly refers to the degree to which a pair of variables are] **linearly** .monash-gray10[related]. [Wikipedia](https://en.wikipedia.org/wiki/Correlation_and_dependence)

<br>
<br>
.info-box[If the .monash-orange2[relationship is not linear], call it .monash-orange2[association], and avoid correlated.]

---
# Peceiving correlation

.panelset[
.panel[.panel-name[🖼️]

.monash-orange2[Let's play a game:] Guess the correlation!

<br>
```{r simcor, fig.width=10, fig.height=5, out.width="90%"}
set.seed(7777)
vc <- matrix(c(1,0,0,1), ncol=2, byrow=T)
d <- as_tibble(rmvnorm(500, sigma=vc))
p1 <- ggplot(d, aes(x=V1, y=V2)) +
  geom_point() + 
  theme_void() +
  theme(aspect.ratio=1, 
        plot.background=element_rect(fill="gray90"))
vc <- matrix(c(1,0.4,0.4,1), ncol=2, byrow=T)
d <- as_tibble(rmvnorm(500, sigma=vc))
p2 <- ggplot(d, aes(x=V1, y=V2)) +
  geom_point() + 
  theme_void() +
  theme(aspect.ratio=1, 
        plot.background=element_rect(fill="gray90"))
vc <- matrix(c(1,0.6,0.6,1), ncol=2, byrow=T)
d <- as_tibble(rmvnorm(500, sigma=vc))
p3 <- ggplot(d, aes(x=V1, y=V2)) +
  geom_point() + 
  theme_void() +
  theme(aspect.ratio=1, 
        plot.background=element_rect(fill="gray90"))
vc <- matrix(c(1,0.8,0.8,1), ncol=2, byrow=T)
d <- as_tibble(rmvnorm(500, sigma=vc))
p4 <- ggplot(d, aes(x=V1, y=V2)) +
  geom_point() + 
  theme_void() +
  theme(aspect.ratio=1, 
        plot.background=element_rect(fill="gray90"))
vc <- matrix(c(1,-0.2,-0.2,1), ncol=2, byrow=T)
d <- as_tibble(rmvnorm(500, sigma=vc))
p5 <- ggplot(d, aes(x=V1, y=V2)) +
  geom_point() + 
  theme_void() +
  theme(aspect.ratio=1, 
        plot.background=element_rect(fill="gray90"))
vc <- matrix(c(1,-0.5,-0.5,1), ncol=2, byrow=T)
d <- as_tibble(rmvnorm(500, sigma=vc))
p6 <- ggplot(d, aes(x=V1, y=V2)) +
  geom_point() + 
  theme_void() +
  theme(aspect.ratio=1, 
        plot.background=element_rect(fill="gray90"))
vc <- matrix(c(1,-0.7,-0.7,1), ncol=2, byrow=T)
d <- as_tibble(rmvnorm(500, sigma=vc))
p7 <- ggplot(d, aes(x=V1, y=V2)) +
  geom_point() + 
  theme_void() +
  theme(aspect.ratio=1, 
        plot.background=element_rect(fill="gray90"))
vc <- matrix(c(1,-0.9,-0.9,1), ncol=2, byrow=T)
d <- as_tibble(rmvnorm(500, sigma=vc))
p8 <- ggplot(d, aes(x=V1, y=V2)) +
  geom_point() + 
  theme_void() +
  theme(aspect.ratio=1, 
        plot.background=element_rect(fill="gray90"))
grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, ncol=4)
```
]
.panel[.panel-name[answers]
.grid4x2[
.item[
<br>
 $\rho=0$ 
<br>
]
.item[
<br>
 $\rho=0.4$ 
<br>
]
.item[
<br>
 $\rho=0.6$ 
<br>
]
.item[
<br>
 $\rho=0.8$ 
<br>
]
.item[
<br>
 $\rho=-0.2$ 
<br>
]
.item[
<br>
 $\rho=-0.5$ 
<br>
]
.item[
<br>
 $\rho=-0.7$ 
<br>
]
.item[
<br>
 $\rho=-0.9$ 
<br>
]
]

<br>
<br>

My guess is that you didn't do very well. You likely under-stated r, particularly around 0.4-0.7. The variation in scatter is not linear with r.

When someone says .monash-blue2[*correlation is 0.5* it sounds impressive]. BUT when someone shows you a .monash-blue2[scatterplot of data that has correlation 0.5], you will say that's a .monash-blue2[weak relationship.]
]
.panel[.panel-name[R]
.scroll-800[
```{r ref.label="simcor", echo=TRUE, eval=FALSE}
```
]
]
]

---
class: transition middle

# Transformations

for skewness, heteroskedasticity and linearising relationships

---
#  .orange[Case study] .bg-orange.circle[2] Movies

.panelset[
.panel[.panel-name[🖼️]
.grid[
.item[
```{r movies, fig.width=6, fig.height=6, out.width="90%"}
ggplot(movies, aes(x=votes, y=rating)) +
  geom_point() + 
  ylim(c(1,10)) #<<
```
]
.item[
- `votes`: Number of IMDB users who rated this movie
- `rating`: Average IMDB user rating

<br>
<br>
Describe the relationship between rating and votes. 
]
]
]
.panel[.panel-name[learn]
.grid[
.item[
- Odd pattern, almost looks like an "r"
- No films with lots of votes and low rating
- No film with lots of votes has rating close to maximum possible: **barrier?**
- Films with very high ratings only have a few votes
- Generally, rating appears to increase as votes increases (its hard to really read this with so few points though)
- A few films with really large number of votes: **outliers?** or just **skewness?**
- Films with few votes have ratings that span the range of the scale.
]
.item[
Would you say this is positive, linear, moderate?

Or positive, non-linear, and moderate? Or weak?

In some sense, these descriptions are meaningless, here.

<br>
What about causation? association? outliers?  clusters? gaps? barrier? conditional relationships?

<br>

.monash-blue2[These descriptive help to describe relationships generally, but it is important to convert them into the context of the (variables in the) data.]

.monash-orange2[BUT, BUT there is a skewness in votes that needs fixing before assessing the relationship.]
]
]
]
.panel[.panel-name[R]

```{r ref.label="movies", echo=TRUE, eval=FALSE}
```
]
]

---
#  .orange[Case study] .bg-orange.circle[2] Movies

.panelset[
.panel[.panel-name[🖼️]
.grid[
.item[
```{r logmovies, fig.width=6, fig.height=6, out.width="90%"}
ggplot(movies, aes(x=votes, y=rating)) +
  geom_point(alpha=0.1) + 
  geom_smooth(se=F, colour="orange", size=2) +
  scale_x_log10() +
  ylim(c(1,10)) #<<
```
]
.item[
<br>
<br>
<br>
<br>
`r emo::ji("thinking")` Something funny happens, right at 1000 votes
<br>
<br>

Some positive association between two variables only for large number of votes.
]
]
]
.panel[.panel-name[R]

```{r ref.label="logmovies", echo=TRUE, eval=FALSE}
```

*Note*: Used .monash-orange2[transparency] (because there is a lot of data) and a .monash-orange2[loess smooth] (because I am interested in assessing the trend between votes and rating).

<br>

Correlation between .monash-blue2[raw variables] is `r round(cor(movies$votes, movies$rating), 2)` <br> and between .monash-blue2[transformed] `log(votes)`and `rating` is `r round(cor(log10(movies$votes), movies$rating), 2)`. Which more accurately reflects the relationship?

What do you expect that removing the outlier will do to the correlation value?
]
]

---
#  .orange[Case study] .bg-orange.circle[3] Cars

.panelset[
.panel[.panel-name[🖼️]
.grid[
.item[
```{r cars, fig.width=6, fig.height=6, out.width="90%"}
data(mtcars)
ggplot(mtcars, aes(x=hp, y=mpg)) +
  geom_point() + 
  geom_smooth(colour="forestgreen", se=F) 
```
]
.item[
- `mpg`: Miles/(US) gallon
- `hp`: Gross horsepower

<br>
<br>
Describe the relationship between horsepower and mpg. 
]
]
]
.panel[.panel-name[learn]

<br>
<br>

- negative: as horsepower increases fuel efficiency is worse
- nonlinear: for lower horse power the decrease in efficieny is more
- strong: very little variation between cars, looks fundamentally like a physics problem
- outlier: one car with high horse power has unusually high efficiency

]
.panel[.panel-name[R]

```{r ref.label="cars", echo=TRUE, eval=FALSE}
```
]
]

---
#  .orange[Case study] .bg-orange.circle[3] Cars

.panelset[
.panel[.panel-name[🖼️]
.grid[
.item[
```{r logcars, fig.width=6, fig.height=6, out.width="90%"}
ggplot(mtcars, aes(x=hp, y=mpg)) +
  geom_point() + 
  scale_y_log10("log mpg") + #<<
  geom_smooth(method="lm", colour="forestgreen", se=F) +
  geom_smooth(data=filter(mtcars, hp<300), method="lm", colour="orangered", se=F, lty=2)
```
]
.item[
- `mpg`: Miles/(US) gallon
- `hp`: Gross horsepower

<br>
<br>
Log transforming `mpg` linearised the relationship between horsepower and mpg. 

<br>
<br>

.monash-orange2[Need to also remove the outlier, because it is clearly influential (swinging the line towards it).]
]
]
]
.panel[.panel-name[R]

```{r ref.label="logcars", echo=TRUE, eval=FALSE}
```


Correlation between .monash-blue2[raw variables] is `r round(cor(mtcars$hp, mtcars$mpg), 2)` <br> and between .monash-blue2[transformed] `log(mpg)` and `hp` is `r round(cor(log10(mtcars$hp), mtcars$mpg), 2)`. Which more accurately reflects the relationship?

]
]
---
# Circle of transformations for linearising

.grid[
.item[ 
```{r circleoftrans, fig.width=5, fig.height=5, out.width="80%"}
x <- c(seq(-0.95, -0.1, 0.025), seq(0.1, 0.95, 0.025))
x <- c(x, x)
d <- tibble(x, y = sqrt(1-x^2), 
            q=c(rep("4", 35), 
                rep("1", 35),
                rep("3", 35),
                rep("2", 35)))
d$y[71:140] <- -d$y[71:140]
ggplot(d, aes(x=x, y=y, colour=q, group=q)) +
  geom_line(size=2) +
  annotate("text", x=c(0.5, 0.5, -0.5, -0.5),
           y=c(0.5, -0.5, -0.5, 0.5), 
           label=c("x up, y up", "x up, y down", 
                   "x down, y down", "x down, y up"),
           size=5) +
  geom_vline(xintercept=0) + geom_hline(yintercept=0) +
  annotate("text", x=0, y=0, label="(0,0)", size=10) +
  theme_void() +
  theme(aspect.ratio=1, legend.position = "none")
```
]
.item[
Remember the power ladder: 

-1, 0, 1/3, 1/2, .monash-orange2[1], 2, 3, 4 ?

<br>

1.Look at the shape of the relationship. 
2.Imagine this to be a number plane, and depending on which quadrant the shape falls in, you either transform $x$ or $y$, up or down the ladder: `+,+` both up; `+,-` x up, y down; `-,-` both down;  `-,+` x down, y up
    
<br>

If there is heteroskedasticity, try transforming $y$, may or may not help

]
]

---
#  .orange[Case study] .bg-orange.circle[4] Soils

Interplay between skewness and association

.grid[
.item[
```{r baker, fig.width=5, fig.height=5, out.width="80%"}
baker <- read_csv("../data/baker.csv")
p <- ggplot(baker, aes(x=B, y=Corn97BU)) +
  geom_point() +
  xlab("Boron (ppm)") +
  ylab("Corn Yield (bushells)") 
ggMarginal(p, type="density")
```

]
.item[
Soil chemical analysis of a farm field in Iowa

<br>

Is there a relationship between Yield and Boron?

<br>
<br>

`ggMarginal` adds density plots for each variable to the scatterplot

<br>
Is Boron skewed? Is Yield skewed? Then it is .monash-orange2[hard] to assess the relationship.
]
]
---
#  .orange[Case study] .bg-orange.circle[4] Soils

.grid[
.item[
```{r transfbaker, fig.width=5, fig.height=5, out.width="80%"}
p <- ggplot(baker, aes(x=B, y=Corn97BU^2)) + #<<
  geom_point() +
  xlab("log Boron (ppm)") +
  ylab("Corn Yield^2 (bushells)") +
  scale_x_log10() #<<
ggMarginal(p, type="density") #<<
```
]
.item[
.font_small[
<br>
<br>

```{r ref.label="transfbaker", echo=TRUE, eval=FALSE}
```
]
]
]

---
#  .orange[Case study] .bg-orange.circle[4] Soils


.grid[
.item[
```{r bakeriron, fig.width=5, fig.height=5, out.width="80%"}
p <- ggplot(baker, aes(x=Fe, y=Corn97BU^2)) + 
  geom_density2d(colour="orange") +
  geom_point() +
  xlab("Iron (ppm)") +#<<
  ylab("Corn Yield^2 (bushells)") 
ggMarginal(p, type="density") 
```
]
.item[
.font_small[
<br>
Lurking variable?

<br>
<br>

```{r ref.label="bakeriron", echo=TRUE, eval=FALSE}
```
]
]
]

---
#  .orange[Case study] .bg-orange.circle[4] Soils

.grid[
.item[
```{r bakerironca, fig.width=5, fig.height=5, out.width="80%"}
ggplot(baker, aes(x=Fe, y=Corn97BU^2,
       colour=ifelse(Ca>5200, "high", "low"))) + #<<
  geom_point() +
  xlab("Iron (ppm)") +
  ylab("Corn Yield^2 (bushells)") +
  scale_colour_brewer("", palette="Dark2") +
  theme(aspect.ratio=1, 
        legend.position = "bottom")
```
]
.item[
.font_small[

<br>
Colour high (>5200ppm) calcium values

<br>
<br>

```{r ref.label="bakerironca", echo=TRUE, eval=FALSE}
```

<br>

If calcium levels in the soil are high, yield is consistently high. If calcium levels are low, then there is a positive relationship between yield and iron, with higher iron leading to higher yields.

]
]
]
---
#  .orange[Case study] .bg-orange.circle[5] COVID-19


.panelset[
.panel[.panel-name[🖼️]

```{r prepdata, eval=FALSE}
# Read data
nyt_county <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")

# Add centroid lat/long
geoCounty_new <- geoCounty %>% as_tibble() %>%
  mutate(fips = as.character(fips),
         county = as.character(county),
         state = as.character(state)) %>%
  select(fips, lon, lat)
nyt_county_total <- nyt_county %>% 
  group_by(fips) %>%
  filter(date == max(date)) %>%
  left_join(geoCounty_new, by=c("fips"="fips"))

save(nyt_county_total, file="../data/nyt_covid.rda")
```

```{r usacovid, fig.width = 10, fig.height = 6.7, out.width="70%"}
load("../data/nyt_covid.rda")
usa <- map_data("state")
ggplot() +
  geom_polygon(data=usa, 
               aes(x=long, y=lat, group=group), 
               fill="grey90", colour="white") +
  geom_point(data=nyt_county_total, 
             aes(x=lon, y=lat, size=cases), 
             colour="red", shape=1) +
  geom_point(data=nyt_county_total, 
             aes(x=lon, y=lat, size=cases), 
             colour="red", fill="red", alpha=0.1, shape=16) +
  scale_size("", range=c(1, 30)) + 
  theme_map() + theme(legend.position="none")
```

Bubble plots, size of point is mapped to another variable.

This bubble plot here shows total count of COVID-19 incidence (as of Aug 30, 2020) for every county in the USA, inspired by the [New York Times coverage](https://www.nytimes.com/news-event/coronavirus).

]
.panel[.panel-name[R]

```{r ref.label="usacovid", echo=TRUE, eval=FALSE}
```
]
]

---
# Scales matter

.grid[
.item[
<br>
<br>

```{r ref.label="usacovid", eval=TRUE, fig.width = 10, fig.height = 6.7, out.width="70%"}
```

]
.item[
<br>
<br>
Where has COVID-19 hit the hardest?
<br>

Where are there more people?
<br>
<br>
<br>

Is it only a problem in population centres? Should we .monash-orange2[calibrate the counts by population]?
]
]

---
# Generalised

> What do you do if the variables are not continuous/quantitative?

The type of variable determines the choice of mapping.

- Continuous and categorical $\longrightarrow$ side-by-side boxplots, side-by-side density plots
- Both categorical $\longrightarrow$ faceted bar charts, stacked bar charts, mosaic plots, double decker plots

<br> <br>

> We'll see more examples soon.
---
# Simpsons paradox

There is an additional variable, which if used for conditioning, changes the association between the variables, you have a .monash-orange2[paradox] `r emo::ji("upside_down_face")`.

```{r generate_data}
set.seed(2222)
df <- tibble(x=c(rnorm(500)*0.2, runif(300)+1)) %>%
  mutate(y1=c(-2*x[1:500]+rnorm(500), 
              3*x[501:800] + rexp(300)), 
         y2=c(rep("A", 500), rep("B", 300)))
```

.grid[
.item[
```{r scat, fig.width=3, fig.height=3, out.width="70%"}
ggplot(df, aes(x=x, y=y1)) +
  geom_point(alpha=0.5) + xlab("") + ylab("") +
  annotate("text", x=0, y=8, label=paste0("r = ", round(cor(df$x,df$y1), 2))) +
  theme_minimal() 
```

]
.item[
```{r scatcol, fig.width=3, fig.height=3, out.width="70%"}
ggplot(df, aes(x=x, y=y1, colour=y2)) +
  geom_point(alpha=0.5) + xlab("") + ylab("") +
  scale_colour_brewer("", palette="Dark2") +
  annotate("text", x=0, y=8, label=paste0("r = ", round(cor(df$x[df$y2=="A"],df$y1[df$y2=="A"]), 2)), colour="forestgreen") +
    annotate("text", x=1.5, y=0, label=paste0("r = ", round(cor(df$x[df$y2=="B"],df$y1[df$y2=="B"]), 2)), colour="orangered") +
  theme_minimal() + theme(legend.position="none")
```

]

]


---
# Simpsons paradox: famous example

<br>
```{r berkeley, fig.width=10, fig.height=4, out.width="90%"}
ucba <- as_tibble(UCBAdmissions)
a <- ggplot(ucba, aes(Dept)) +
  geom_bar(aes(weight=n))
b <- ggplot(ucba, aes(Admit)) +
  geom_bar(aes(weight=n)) +
  facet_wrap(~Gender)
grid.arrange(a, b, ncol=2)
```

<br>

Did Berkeley .monash-orange2[discriminate] against female applicants?

.footnote[Example from Unwin (2015)]

---
# Simpsons paradox: famous example

```{r berkeleydd, fig.width=10, fig.height=4, out.width="100%"}
library(vcd)
ucba <- ucba %>%
  mutate(Admit = factor(Admit, 
            levels=c("Rejected", "Admitted")),
         Gender = factor(Gender, 
            levels=c("Male", "Female"),
            labels=c("M","F")))
doubledecker(xtabs(n~Dept+Gender+Admit, data=ucba),
             gp = gpar(fill=c("grey90", "orangered")))
```

Based on separately examining each department, there is .monash-orange2[no evidence of discrimination] against female applicants.

.footnote[Example from Unwin (2015)]
---
# Spurious association

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">This graph is amazing. It shows that measuring <a href="https://twitter.com/hashtag/SARSCoV2?src=hash&amp;ref_src=twsrc%5Etfw">#SARSCoV2</a> levels in municipal sewage almost perfectly predicts forthcoming <a href="https://twitter.com/hashtag/COVID19?src=hash&amp;ref_src=twsrc%5Etfw">#COVID19</a> cases with a full week&#39;s notice (R=0.994). It&#39;s one of several discoveries in this new study from <a href="https://twitter.com/Yale?ref_src=twsrc%5Etfw">@Yale</a>: <a href="https://t.co/hZVdXebx2D">https://t.co/hZVdXebx2D</a>. C-19 is <a href="https://twitter.com/hashtag/InThePoop?src=hash&amp;ref_src=twsrc%5Etfw">#InThePoop</a> <a href="https://t.co/shNzTKYoYP">pic.twitter.com/shNzTKYoYP</a></p>&mdash; Brennan Spiegel, MD, MSHS (@BrennanSpiegel) <a href="https://twitter.com/BrennanSpiegel/status/1265119535901732865?ref_src=twsrc%5Etfw">May 26, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

---
count: false
# Spurious association

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">This preprint is getting a lot of likes and retweets. But a correlation of .994 when one of the variables is an integer in the range (0,29) seems... optimistic. /1 <a href="https://t.co/fmGJEhy0sO">https://t.co/fmGJEhy0sO</a></p>&mdash; Nick Brown (@sTeamTraen) <a href="https://twitter.com/sTeamTraen/status/1265411882283917315?ref_src=twsrc%5Etfw">May 26, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

---
count: false
# Spurious association

.scroll-800[
<blockquote class="twitter-tweet"><p lang="en" dir="ltr">The data don&#39;t seem to be readily available, so I digitised them by eye from Figure 2A. This will certainly be imperfect, but it&#39;s quite easy to calibrate the red [virus] dots, since the grey [admissions] dots must be integers. /2 <a href="https://t.co/CYPoZcTygN">pic.twitter.com/CYPoZcTygN</a></p>&mdash; Nick Brown (@sTeamTraen) <a href="https://twitter.com/sTeamTraen/status/1265411885324800007?ref_src=twsrc%5Etfw">May 26, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
]

---
count: false
# Spurious association

.scroll-800[
<blockquote class="twitter-tweet"><p lang="en" dir="ltr">More exact raw data rescaled to [0, 1] as extracted from the PDF: <a href="https://t.co/xrjMisMilT">pic.twitter.com/xrjMisMilT</a></p>&mdash; Funko Unko (@FunkoUnko) <a href="https://twitter.com/FunkoUnko/status/1265571596657442818?ref_src=twsrc%5Etfw">May 27, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
]
---
class: informative middle
count: false
# Spurious association

.info-box[Be careful of double time series, especially if the data is not being shown.]

---
# Checking association with randomisation

.panelset[
.panel[.panel-name[Soils]

```{r soils-lineup, fig.width=10, fig.height=7, out.width="80%"}
ggplot(lineup(null_permute("Corn97BU"), baker, n=12), 
       aes(x=B, y=Corn97BU)) +
  geom_point() +
  facet_wrap(~.sample, ncol=4)
```

]
.panel[.panel-name[R]
```{r ref.label="soils-lineup", echo=TRUE, eval=FALSE}
```

11 of the panels have had the association broken by permuting one variable. .monash-blue2[There is no association] in these data sets, and hence plots. Does the data plot stand out as being different from the null (no association) plots?

]
.panel[.panel-name[Olympics]

```{r oly-lineup, fig.width=10, fig.height=7, out.width="80%"}
data(oly12, package = "VGAMdata") 
oly12_sub <- oly12 %>%
  filter(Sport %in% c("Swimming", "Archery",
                      "Hockey", "Tennis")) %>%
  filter(Sex == "F") %>%
  mutate(Sport = fct_drop(Sport), Sex=fct_drop(Sex)) 

ggplot(lineup(null_permute("Sport"), oly12_sub, n=12),
       aes(x=Height, y=Weight, colour=Sport)) + 
  geom_smooth(method="lm", se=FALSE) + 
  scale_colour_brewer("", palette="Dark2") +
  facet_wrap(~.sample, ncol=4) +
  theme(legend.position="none")
```

]
.panel[.panel-name[R]
```{r ref.label="oly-lineup", echo=TRUE, eval=FALSE}
```

11 of the panels have had the association broken by permuting the Sport label. .monash-blue2[There is no difference in the association between weight and height across sports] in these data sets, and hence plots. Does the data plot stand out as being different from the null (no association difference between sports) plots?

]
]

---
# Resources

- Friendly and Denis "Milestones in History of Thematic Cartography, Statistical Graphics and Data Visualisation" available at http://www.datavis.ca/milestones/
- Unwin (2015) [Graphical Data Analysis with R](http://www.gradaanwr.net)
- Graphics using [ggplot2](https://ggplot2.tidyverse.org)



---

```{r endslide, child="components/endslide.Rmd"}
```
